Modelling the gas kinematics of an atypical Lyman alpha emitting compact dwarf galaxy

### 2017-06-23 paper/paper.tex
consistent picture

\begin{abstract}
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe and
  are the best systems to test models on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad, single peaked, highly symmetric \lya\ line emission that had
evaded theoretical interpretation so far.  
In this paper we explain these features by two different physically
motivated kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these kinematic conditions.
We find that both models independently require high velocities, either
a clump velocity dispersion of \sigmaclump\ with outflows of
\inftyclump\ or a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$. 
The velocity dispersion in the multiphase model is consistent with a dynamical
mass of $2\times 10^{9}$ M$_{\odot}$, ten times its baryonic mass,
which can be explained if \tol\ is hosted by a dark matter halo of at least
$10^{11}$ M$_{\odot}$. 
These conditions are similar to Dwarf Elliptical Galaxies.
This work shows the importance of considering multiphase
and rotation kinematics among the possible conditions shaping the
\lya\ spectra.
If future kinematic maps of \tol\ confirm the velocities found by the
multiphase model, it will be the basis to study new processes that
might be at play during priomordial galaxy evolution.
\end{abstract}

@@

\tol\ is a compact star forming dwarf galaxy that presents a
strong \lya\ emission with two puzzling 
features: the line is highly symmetric, single peaked and wide around the
line's center \citep{Thuan97}.
This kind of profile also raises the question whether some high
redshift LAEs have asymmetric lines because the blue half was
truncated by the intergalactic medium \citep{2007MNRAS.377.1175D}. 
In this case the \lya\ radiation could emerge as a low surface
brightness glow, which may be connected to \lya\ halos, while also
influencing the way LAEs can be used as a probe of reionization
See the review by \cite{2014PASA...31...40D} and the references therein.

@@

The main motivation for the outflowing multiphase model (as presented
in \cite{Gronke2016}) is that some dwarf galaxies are expected to
present outflows.  
Observationally, outflows have been detected in few local dwarf galaxies
\citep{1998ApJ...506..222M,2005MNRAS.358.1453O}. 
Clumpy multiphase outflows are capable to explain \lya\ features around
star-forming galaxies
\citep{2010ApJ...717..289S,2012MNRAS.424.1672D}. 
From this evidence we decide to try a multiphase model.

The motivation for the rotation without outflows (as presented in \cite{GaravitoCamargo2014})is that
dwarf galaxies show coherent rotation features
\citep{2009A&A...493..871S} and it is expected that some of them have
high neutral gas contents with long quiescent phases without high gas
outflows triggered by supernova activity
\citep{2005A&A...433L...1B,2008ApJ...672..888T,2013MNRAS.434.2491G}.  
For these reasons we also explore a model with a spherical gas
distribution with solid body rotation and without any outflows.

@@

We do not claim that our findings discard the existence of different
undiscovered models that could reproduce the observed features in \tol.
More involved models with multiple kinematic components could
in principple be also sucessful.
We focus on the existence of two simple and succesful models that
manage, for the first time, to account for \tol's \lya\ line main
features.

We review first the observational characteristics of
\tol, then we summarize the main features in the multiphase and
rotation models to explain how we fit their free parameters 
to the \tol's \lya\ line shape.
We use the results to interpret them in terms of the galaxy's
dynamical mass. 
We close by discussing the physical scenarios opened up by our
results and the additional observational information needed to clarify
the kinematic nature of \tol.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.55\textwidth]{CLARA-TOL-main.pdf}
\caption{Broad, single peaked and symmetric \lya\ emission of \tol.
  Dots correspond to the observational data \citep{mashesse03}. 
The thick red and thin blue curves represent our best fit
to the data using the full radiative transfer simulation with a
rotation and multiphase model, respectively. 
These two different models are able to reproduce the main morphological
features of \tol's \lya\ line emission.}
\end{center}
\end{figure*}

@@

The idealized multiphase model consists of spherical, cold, dense
clumps of neutral hydrogen (and dust) embedded in a hot, ionized
medium \citep{Gronke2016}. 
The clumps also have a random and an outflowing velocity
component which totals the number of parameters describing the model
to be $14$, which also include information about the spatial
distribution and thermal state of the clumps and the intraclump
medium.

@@

\begin{equation}
\chi^2 = \sum_{i} \frac{({f}_i - \hat{f}_i)^2}{\sigma_i^2}, 
\label{eq:chi2}
\end{equation}
%
where $i$ iterates over velocity bins, $f_{i}$ is the observed flux,
$\sigma_i$ is the observed flux uncertainty and $\hat{f}_i$ is the
model estimated flux.

We do not have analytic expression for $\hat{f}$. 
For the multi-phase model the results of the MonteCarlo are used to
build histograms that play the role of $\hat{f}$. 
For the rotation model $\hat{f}$ is computed 
has to be computed by numerical integration of analytical
approximations found in  \citet{GaravitoCamargo2014}.

@@

We use an analytical approximation that captures the most important
effects of rotation onto the \lya\ line. 
We defer the reader to \citet{GaravitoCamargo2014} for complete
details on the explicit form of this approximation.
To fully explore the
parameter space using a Markov Chain Monte Carlo (MCMC) calculation with the
\texttt{emcee} Python library \citep{2013PASP..125..306F}. \texttt{emcee} 
is an open source optimized implementation of the affine-invariant 
MCMC sampler \citep{goodman2010ensemble}. 
The algorithm creates a number of walkers that,
during a sufficient number of steps, generate parameters' combinations
for a specific model.
For each timestep, the code calculates the likelihood of the
combination with respect to the observational data.
The walkers explore the parameter space sampling the Gaussian likelihood
function built as $\propto \exp(-\chi^2/2)$, where the $\chi^2$ follows
the definition in Equation \ref{eq:chi2}.

@@

In spite of the simplicity of our models, this is the first time that
the main \tol\ features can be reproduced: a wide, centered,
single-peaked \lya\ line.
This result does not demonstrate that the kinematic features we
include in our models are necessary to reproduce \tol's features, but
at least they show that they are a sufficient condition.
This is a significant step forward to understand the influence that
different kinematics have in producing the atypical line profile shown
by \tol.

@@

Qualitatively this can be understood as follows. The broad \lya\ line
is the result of subsequent \lya\ photon scatterings due to
(relatively) fast moving clumps while the multi-phase nature (i.e., the existence of low-density
inter-clump regions where \lya\ photons can freely prograpagate)
ensures the high flux at line center as observed.  
In turn, a lower velocity dispersion would produce a thiner line than
it is observed.
From the lower right panel in Figure \ref{multiphaseresults} we find that it
is unlikely that the clump velocity dispersion is lower than
$50$\kms.

Why is it that the other parameters do not seem to matter? Probably
because $P_{cl}$ (the probability for a \lya\ to be emitted in a
clump) is high and most photons are produced at
large radii making the radiative transfer effects in the center less
important.  
We cannot discard that another region of parameter space with similar
velocity structure but an empty inter-clump medium, relatively dense
clumps and at least a few clumps per sightline 
could also work \citep{Hansen06}.
This possibility does not change our conclusions about the
kinematic parameters, but it might be interesting to explore in the
future with a higher sampling of the parameter space once new 
observations  clarify the kinematic nature of \tol \footnote{In Section \S \ref{sec:conclusions} we describe  in more detail what kind of observations could be performed in the
  near future to improve our understanding of \tol's kinematics.} .

@@

\tol\ shows one of the most atypical \lya\ spectra observed so far.
Its flux at the line's center is high compared to other LAEs at low
redshift and the wide symmetrical peak is virtually absent from other
LAEs at low and high redshift \citep{2012ApJ...751...29Y,LARS,Erb14,Trainor16}. Simple
shell models fail to reproduce such a spectrum
\cite{2015A&A...578A...7V}.  
In the previous sections we show how these characteristics can be
explained by two different kinematic models. Which model should we prefer?

\tol's structural properties seem to be similar to observed dwarf
elliptical galaxies \citep{2003AJ....126.1794G}. 
Galaxies such as IC 3344 and NGC 4482 have similar absolute $M_V$
magnitudes and size as \tol. 
These galaxies have velocity dispersions in the range $30$-$45$ \kms. 
The rotational velocities of those systems is lower than $1$\kms. 
This supports the idea that the multiphase model the correct model
describing \tol, as it provides at least some consistency with
kinematic properties of other dwarf elliptical galaxies.

>>>>

\begin{abstract}
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe and
  are the best systems to test models on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad, single peaked, highly symmetric \lya\ line emission that had
evaded theoretical interpretation so far.  
In this paper we explain these features by two different physically
motivated kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these kinematic conditions.
We find that both models independently require high velocities, either
a clump velocity dispersion of \sigmaclump\ with outflows of
\inftyclump\ or a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$. 
The velocity dispersion in the multiphase model is consistent with a dynamical
mass of $2\times 10^{9}$ M$_{\odot}$, ten times its baryonic mass,
which can be explained if \tol\ is hosted by a dark matter halo of at least
$10^{11}$ M$_{\odot}$; cconditions similar to those of Dwarf Elliptical Galaxies.
This work shows the importance of considering multiphase
and rotation kinematics among the possible conditions shaping the
\lya\ spectra.
If future kinematic maps of \tol\ confirm the velocities found by the
multiphase model, it will be the basis to study new processes that
might be at play during priomordial galaxy evolution.
\end{abstract}

@@

\tol\ is a compact star forming dwarf galaxy that presents a
strong \lya\ emission with puzzling 
features: the line is highly symmetric, single peaked and broad around the
line's center \citep{Thuan97}.
This kind of profile also raises the question whether some high
redshift LAEs have asymmetric lines because the blue half was
truncated by the intergalactic medium \citep{2007MNRAS.377.1175D}. 
In this case the \lya\ radiation could emerge as a low surface
brightness glow, which may be connected to \lya\ halos, while also
influencing the way LAEs can be used as a probe of reionization
See the review by \cite{2014PASA...31...40D} and the references therein.

@@

The main motivation for the outflowing multiphase model (as presented
in \cite{Gronke2016}) is that some dwarf galaxies are expected to
present outflows.  
Observationally, outflows have been detected in few local dwarf galaxies
\citep{1998ApJ...506..222M,2005MNRAS.358.1453O}. 
Clumpy multiphase outflows are capable to explain \lya\ features around
star-forming galaxies
\citep{2010ApJ...717..289S,2012MNRAS.424.1672D}.

The motivation for the rotation without outflows (as presented in \cite{GaravitoCamargo2014})is that
dwarf galaxies show coherent rotation features
\citep{2009A&A...493..871S} and it is expected that some of them have
high neutral gas contents with long quiescent phases without high gas
outflows triggered by supernova activity
\citep{2005A&A...433L...1B,2008ApJ...672..888T,2013MNRAS.434.2491G}.

@@

We do not claim that our findings discard the existence of different
undiscovered models that could reproduce the observed features in \tol.
More involved models with multiple kinematic components could
in principple be also sucessful.
We focus on presenting two simple and succesful models that
manage, for the first time, to account for \tol's \lya\ line main
features.

We review first the observational characteristics of
\tol, then we summarize the main features in the multiphase and
rotation models to explain how we fit their free parameters 
to the \tol's \lya\ line shape.
We use the results to interpret them in terms of the galaxy's
dynamical mass and explain why the multiphase model should be prefered
over the rotation model.
We close by discussing the physical scenarios opened up by our
results and the additional observational information needed to clarify
the kinematic nature of \tol.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.55\textwidth]{CLARA-TOL-main.pdf}
\caption{Broad, single peaked and symmetric \lya\ emission of \tol.
  Dots correspond to the observational data \citep{mashesse03}. 
The thick red and thin blue curves represent our best fit
to the data using the full radiative transfer simulation with a
rotation and multiphase model, respectively. 
These two different models are able to reproduce the main morphological
features of \tol's \lya\ line emission.\label{fig:results}}
\end{center}
\end{figure*}

@@

The idealized multiphase model consists of spherical, cold, dense
clumps of neutral hydrogen (and dust) embedded in a hot, ionized
medium \citep{Gronke2016}. 
The clumps also have a random and an outflowing velocity
component which totals the number of parameters describing the model
to be $14$, which also include information about the spatial
distribution and thermal state of the clumps and the intraclump
medium. 
We do not explore inflowing clumps given the slight line asymmetry
redwards to the line center (see the dots in Figure \ref{fig:results})

@@

\begin{equation}
\chi^2 = \sum_{i} \frac{({f}_i - \hat{f}_i)^2}{\sigma_i^2}, 
\label{eq:chi2}
\end{equation}
%
where $i$ iterates over velocity bins, $f_{i}$ is the observed flux,
$\sigma_i$ is the observed flux uncertainty and $\hat{f}_i$ is the
model estimated flux.
We do not have analytic expression for $\hat{f}$, instead we use the
results of the MonteCarlo to build histograms that play the role of
$\hat{f}$.

@@

We use an analytical approximation that captures the most important
effects of rotation onto the \lya\ line. 
We defer the reader to \citet{GaravitoCamargo2014} for complete
details on the explicit form of this approximation.
To fully explore the
parameter space using a Markov Chain Monte Carlo (MCMC) calculation with the
\texttt{emcee} Python library \citep{2013PASP..125..306F}. \texttt{emcee} 
is an open source optimized implementation of the affine-invariant 
MCMC sampler \citep{goodman2010ensemble}. 
The algorithm creates a number of walkers that,
during a sufficient number of steps, generate parameters' combinations
for a specific model.
For each timestep, the code calculates the likelihood of the
combination with respect to the observational data.
The walkers explore the parameter space sampling the Gaussian likelihood
function built as $\propto \exp(-\chi^2/2)$, where the $\chi^2$ follows
the definition in Equation \ref{eq:chi2}. 
We do not have an analytic expresion for $\hat{f}$, we compute it
by numerical integration of the analytical approximations found in
\citet{GaravitoCamargo2014}.

@@

The fit to the observations is not perfect. However, in spite of the
simplicity of our models, this is the first time that the main
\tol\ features can be reproduced: a broad, highly symmetric,
single-peaked \lya\ line.

This result does not demonstrate that the kinematic features we
include in our models are necessary to reproduce \tol's features, but
at least they show that they are a sufficient condition.
This is a significant step forward to understand the influence that
different kinematics have in producing the atypical line profile shown
by \tol.

@@

We tested with a set of 100 additional Monte-Carlo simulations
with random values for $\sigma_{\rm cl}$, $v_{\infty {\rm cl}}$ and
$P_{\rm cl}$ within the derived constraints (and random values for the
other parameters over the whole parameter range) and found that $30$
models result in $\chi^2$ values  
continue to provide a similar $\chi^2$ values as the best $25$ models.

Qualitatively this result can be understood as follows. 
The broad \lya\ line is the result of subsequent \lya\ photon
scatterings due to (relatively) fast moving clumps while the
multi-phase nature (i.e., the existence of low-density inter-clump
regions where \lya\ photons can freely prograpagate)  ensures the high
flux at line center as observed.   
In turn, a lower velocity dispersion would produce a thiner line than
it is observed.
From the lower right panel in Figure \ref{multiphaseresults} we find that it
is unlikely that the clump velocity dispersion is lower than
$50$\kms.

Having constrained only 3 parameters, one migh wonder why the other 11
parameters do not seem to matter. 
However, this view is misleading. 
Event if for clumply models the line profile is most sensitive to only
a few parameters \citep{Gronke}, there are the rest of parameters
which affect the line profile at a level that cannot be neglected,
i.e. we cannot randomly pick 11 random, reasonable values for these
parameters, and expect to fit the line profile while only varying the
other 3.

We cannot discard that another region of parameter space with similar
velocity structure but an empty inter-clump medium, relatively dense
clumps and at least a few clumps per sightline 
could also work \citep{Hansen06}.
This possibility does not change our conclusions about the
kinematic parameters, but it might be interesting to explore in the
future with a higher sampling of the parameter space once new 
observations  clarify the kinematic nature of \tol \footnote{In
  Section \S \ref{sec:conclusions} we describe  in more detail what
  kind of observations could be performed in the near future to
  improve our understanding of \tol's kinematics.} .

@@

\tol\ shows one of the most atypical \lya\ spectra observed so far.
Its flux at the line's center is high compared to other LAEs at low
redshift and the broad, highly symmetrical peak is virtually absent from other
LAEs at low and high redshift
\citep{2012ApJ...751...29Y,LARS,Erb14,Trainor16}. Simple shell models
fail to reproduce such a spectrum as reported by \cite{2015A&A...578A...7V}.  
In the previous sections we show how these characteristics can be
explained by two different kinematic models. Which model should we prefer?

\tol's structural properties seem to be similar to observed dwarf
elliptical galaxies \citep{2003AJ....126.1794G}. 
Galaxies such as IC 3344 and NGC 4482 have similar absolute $M_V$
magnitudes and size as \tol. 
These galaxies have velocity dispersions in the range $30$-$45$ \kms
and their rotational velocities are lower than $1$\kms. 
This supports the idea that the multiphase model the correct model
describing \tol, as it provides at least some consistency with
kinematic properties of other dwarf elliptical galaxies.
